Temporal ordering and registration of images in studies of developmental dynamics

### 2015-01-31 image_analysis_paper_development2/manuscript.tex
updated paper

\raisebox{4.25cm}{\textsf{C}}
\includegraphics[width=5cm]{drosophila_live_imaging_angle_corr}
%
%\hfill
%
\raisebox{4.25cm}{\textsf{D}}
\includegraphics[width=5cm]{drosophila_live_imaging_rank_corr}
%
%\hfill
%
\raisebox{4.25cm}{\textsf{E}}
%
\raisebox{2.5cm}{
{ \textsf{ 
\begin{tabular}{ |c | c| c| }
\hline
\parbox[c]{1.5cm}{ \centering Movie Index} & \parbox[c]{1.5cm}{ \vspace{.5\baselineskip} \centering Average Angle Error \vspace{.5\baselineskip}} & \parbox[c]{1.5cm}{ \centering Rank Correlation Coefficient} \\
\hline
1 & 8.37$^{\circ}$ & 0.9989 \\ 
2 & 4.98$^{\circ}$ & 0.9994 \\ 
3 & 6.47$^{\circ}$ & 0.9795 \\ 
4 & 7.03$^{\circ}$ & 0.9953 \\ 
5 & 3.04$^{\circ}$ & 0.9959 \\ 
 \hline
\end{tabular}}}}
\caption{Method validation using live imaging of \textit{Drosophila} embryos. (A) Selected images from a live imaging study of a \textit{Drosophilla} embryo during gastrulation. Each frame is in an arbitrary rotational orientation, and the order of the frames has been shuffled. (B) Images from A after registration and ordering using vector diffusion maps. The dorsal side of each embryo now appears at the top of each image, and the ventral side appears at the bottom. (C) The correlation between the recovered rotation angle (using vector diffusion maps) and the true rotation angle. (D) The correlation between the recovered rank (using vector diffusion maps) and the true rank. (E) The average error in the recovered angle and the rank correlation coefficient for $5$ different live imaging studies. }
\label{fig:drosophila_live_imaging}
\end{figure*}

@@

We used vector diffusion maps to register and order the images. 
%
\fig~\ref{fig:drosophila_live_imaging}B shows the images from \fig~\ref{fig:drosophila_live_imaging}A, now registered and ordered; the real time for each frame is also indicated.
%
With a small number of exceptions, the recovered ordering is consistent with the real time dynamics. 
%
\fig~\ref{fig:drosophila_live_imaging}C and \fig~\ref{fig:drosophila_live_imaging}D  show the correlations between the recovered and true angles and ranks, respectively, for the entire data set: 
%
Both the angles and the ranks are recovered with a high degree of accuracy.

@@

In general, to accurately recover the temporal order, the vector diffusion maps algorithm requires a sufficient amount of data so that the underlying one-dimensional curve (i.e., the curve in \fig~\ref{fig:schematics}C) is well-sampled.
%
At the same time, the accuracy of the recovered alignments is a function of the signal-to-noise ratio in the pairwise alignments, as well as the size of the data set \citep{singer2011angular}.
%
%Clearly, $40$ images are sufficient to accurately recover the rotational orientation and temporal order in this specific developmental time frame.
%%
%To quantitatively assess how our methods perform as a function of sample size, we used bootstrap sampling to compute the rank correlation as a function of the size of the data set.
%%
%Subsets of images were randomly selected from the $40$-image data set, and then registered and ordered. 
%%
%\fig~\ref{subfig:movie_bootstrap} shows the median rank correlation coefficient as a function of the number of images in the subset.
%%
%As expected, the accuracy in the recovered ordering increases with the number of images. 
%% 
%At the same time, we see that correct order is recovered even with a relatively small number of snapshots, which should make our approach applicable other studies of developmental processes.
%%
Because our particular data set contains relatively low noise levels, few images were required to obtain accurate rotational alignments and the errors in the recovered angles were consistently less than $10^{\circ}$.
%
In summary, the algorithm accurately and robustly registers and orders a model data set where both the rotation angles and real times are independently known.

@@

\vspace{0cm}
\raisebox{2.5cm}{\textsf{C}}
%
\includegraphics[width=3.25cm]{zebrafish_rank_corr}
\end{minipage}
\caption{Method validation using live imaging of a zebrafish embryo. (A) Selected images from a time-lapse movie of zebrafish embryogenesis. Each frame is in an arbitrary rotational orientation, and the order of the frames has been shuffled. (B) Images from A after registration and ordering using vector diffusion maps. The real time of each frame is also indicated. (C) Correlation between the rank recovered using vector diffusion maps and the true rank.  The rank correlation coefficient is $0.9954$.}
\label{fig:zebrafish}
\end{figure*}

To illustrate the broad applicability of our approach, we also applied our algorithm to a time-lapse movie  of zebrafish embryogenesis. 
%
We used a publicly available flipbook movie of zebrafish embryogenesis, available at \texttt{https://zfin.org/zf\_info/movies/Zebrafish.mov} \cite{karlstrom1996flipbook}.  
%
Taken with a differential interference contrast (DIC) microscope, the movie records the first $17$~hours of zebrafish development, from a single cell stage to a 16-somite stage. 
%
We selected $120$ frames from this movie that captures 5.5~hours of epiboly (3.5--9~hours after fertilization). 
%
At the start of the time window, cells have divided 10--11 times and are accumulated in a cell mass above the yolk. 
%
The cell mass is then compressed and the animal-vegetal axis of the embryo (top/bottom in the figure) shortens to form a spherical embryo shape by the end of the fourth hour of development. 
%
Then, the yolk syncytial layer, which forms the boundary between the yolk and the cell mass, moves upward, forming a dome-shaped structure.  
%
During this stage, the cells rearrange to form a cell layer with uniform thickness (about 4-cells thick). 
%
With time, this cell layer then spreads over across the yolk and expands toward the vegetal pole. 
%
At the end of epiboly, the blastoderm completely engulfs the yolk.

As in the example of {\em Drosophila} embryo live imaging, the frames were randomly rotated and shuffled (\fig~\ref{fig:zebrafish}A). 
%
We then used vector diffusion maps to register and order the frames. 
%
The results are shown in \fig~\ref{fig:zebrafish}B. 
%
The recovered rotations and order are consistent with the expected developmental dynamics, as shown in the correlations between the recovered and true ranks (\fig~\ref{fig:zebrafish}C). 
%
Some mismatch in early embryos result from rather static cell movement during early developmental stage where cells divide and pile up above the yolk. 
%
During epiboly, cell movement is more dynamic and the recovered ordering is more consistent with the real dynamics.
%
Quantitatively, the rank correlation coefficient for this data set is 0.9954, and the average error in the recovered angle is $4.14^\circ$.

In summary, we showed that our algorithm is not limited to a specific set of data. 
%
The algorithm works for different tissue types ({\em Drosophila} and zebrafish), and also for images taken with different imaging methods (fluorescent and phase contrast images).
%
Provided there exist significant dynamics within the data set, the developmental dynamics can be recovered.

@@

We have analyzed how our algorithm performs on two model data sets, where all images come from a single embryo. 
%
In practice, each image comes from a different embryo, and the largest source of noise in the considered data set arises from embryo-to-embryo variability.
%
To demonstrate that our methods are robust to such noise sources,
we constructed a synthetic time course data set by selecting an image from a random {\em Drosophila} live imaging experiment (those experiments used in \fig~\ref{fig:drosophila_live_imaging}) at each time point.
%
The resulting data set is spatially unregistered, scrambled in time, and reflects embryo-to-embryo variability. 
%
The median rank correlation coefficient when ordering such a synthetic time course using our methodology was 0.77, indicating that the algorithm can accurately recover the temporal order even under noisy conditions.

@@

\fig~\ref{fig:drosophila_fixed_images}B shows the selected images in \fig~\ref{fig:drosophila_fixed_images}A, now registered and ordered using vector diffusion maps \citep{singer2012vector}.
%
However, because of interembryo variability, these selected images are not entirely representative of the developmental dynamics across the entire data set.
%
To highlight the developmental dynamics revealed by vector diffusion maps, we computed an average developmental trajectory across the set of registered and ordered images (see \SI). 
%
Each snapshot in the average trajectory is the (weighted) average of a group of successive images from the registered and ordered data set.
%
Sequential snapshots of this averaged trajectory, shown in \fig~\ref{fig:drosophila_fixed_images}C, serve as a summary of the stereotypic developmental dynamics.

@@

We have demonstrated that our algorithm can recover the underlying developmental dynamics in several different types of imaging data sets, including ones containing interembryo variability. 
%
Thus far, we have only looked at two-dimensional data sets where biological samples are imaged as a single confocal slice or as a phase contrast image. 
%
However, many developmental processes are inherently three-dimensional, and the z-axis often contains interesting dynamics which are independent of the dynamics in the x-y plane. 
%
Researchers often take multiple z-slices to visualize the three-dimensional structure of a sample;
with advances in imaging techniques, such three-dimensional data will be more common, and it is important to validtate that our algorithm can be applied to such data.

We used an existing three-dimensional data set of fixed {\em Drosophila} wing imaginal discs, which was published in \citet{hamaratoglu2011dpp}.  
%
Imaginal discs are groups of progenitor cells in fly larva that will transform into a specific organ, such as wings, legs, and eyes, in the adult fly during the pupal transformation. 
%
The wing disc is an imaginal disc that turns into a fly wing. 
%
The data set is composed of 46 fixed wing discs, whose developmental time range from 72 to 112~hours after fertilization. 
%
Each disc contains 21 z-slices taken at 1~$\mu$m intervals. 
%
The discs were stained with antibodies that recognize Spalt (red), Dad-GFP (green), and Wingless/Patched (gray) (\fig~\ref{fig:wing_disc}A). 
%
The disc grows in time; at long times, the disc becomes too large to fit completely within the frame and only a partial disc is imaged.

Although the theory of vector diffusion maps readily extends to three-dimensional data, calculation of the pairwise alignments rapidly becomes computational prohibitive for three dimensional data, as one must now search over all three-dimensional rotations (which are parameterized by three Euler angles), rather than over two-dimensional rotations (which are parameterized by a {\em single} rotation angle). 
%
However, ordering data via diffusion maps is readily possible for (registered) three-dimensional data. 
%
Therefore, we can apply our methodology to three-dimensional data by first registering the data using an alternative methodology (such as registering maximum intensity projection, or aligning a single feature of the data), and then ordering the registered data using diffusion maps. 
%
Instead, we used maximum intensity projection and converted the image into 2D, and then used the angular synchronization on 2D data. 
%
For the wing disc data, we first registered the three-dimensional data by computing the optimal rotations to align the maximum intensity projections of the data using vector diffusion maps.
%
Because the maximum intensity projections are two-dimensional images, this step is no more computationally intensive than the previous examples. 
%
Other options for registering samples include using a specific localized signal (such as Twi in \fig~\ref{fig:drosophila_fixed_images}) as a marker, or using other {\em a priori} knowledge.
%
We then used diffusion maps to order the registered three-dimensional images. 
%
\fig~\ref{fig:wing_disc}B shows selected images from the data set ordered by diffusion maps. 
%
In the original data set, each disc was assigned to 6 time classes (72--73~hr, 76.5--77.5~hr, 79--80~hr, 89--90~hr, 100--101~hr, and 110.5--111.5~hr after fertilization); these times are indicated in \fig~\ref{fig:wing_disc}B. 
%
In the ordered set, the size of wing disc grows, and the intensity of proteins increases as a function of time. 
%
The rank correlation coefficient based on the time class is 0.9427. 
%
Furthermore, we can extract a stereotypical developmental trajectory by averaging; this trajectory is shown in \fig~\ref{fig:wing_disc}.
%
Thus, this result shows that the algorithm can successfully order a three-dimensional data set.

@@

\raisebox{1.75cm}{\textsf{D}}
{\renewcommand{\arraystretch}{1.5}
{ \textsf{ \small 
\begin{tabular}{ | p{2cm} | c | c | c | c | c | c | }
\hline
Data Set & 
Data Type & 
\parbox[c]{2cm}{ \centering Number of Channels} & 
\parbox[c]{2cm}{ \centering \vspace{.5\baselineskip} Number of Images \vspace{.5\baselineskip}} & 
\parbox[c]{2cm}{ \centering Number of Pixels} & 
\parbox[c]{2cm}{ \centering Angular Discretization} & 
\parbox[c]{2cm}{ \centering CPU time}  \\
\hline
{\em Drosophila} live~imaging & 2D & 1 & 40 & 100 $\times$ 100 & 10$^\circ$ & 3.2~sec \\ 
Zebrafish & 2D & 1 & 120 & 100 $\times$ 100 & 10$^\circ$ & 13~sec\\ 
{\em Drosophila} fixed images & 2D & 3 & 120 & 100 $\times$ 100 & 10$^\circ$ & 29~sec  \\ 
Wing disc & 3D & 3 & 46 & 100 $\times$ 100 $\times$ 21 & 10$^\circ$ & 12~sec  \\
 \hline
\end{tabular}}}}
\caption{Computational requirements for the presented methodology. (A) CPU time as a function of the number of images in the data set (for $100 \times 100$ pixel images, and $10^\circ$ rotational discretizations). Empirically, the CPU time is $\sim \mathcal{O}(n^{1.33})$ in number of images. (B) CPU time as a function of the number of pixels in the images (for $120$ images, and $10^\circ$ rotational discretizations). Empirically, the CPU time is $\sim \mathcal{O}(n^{1.83})$ in number of pixels. (C) CPU time as a function of the number of rotations (for $120$ images of $100 \times 100$ pixels). Empirically, the CPU time is $\sim \mathcal{O}(n^{-0.77})$ in the angular discretization. (D) The algorithm settings and computational requirements for the data sets analyzed. All times are reported for an Intel Core i7 2.93 GHz processor.}
\label{fig:cpu_time}
\end{figure*}

Manually ordering the images is nontrivial for researchers who are unfamiliar with the developmental progression, and can be tedious and time-consuming for those who are.
%
In contrast, our method requires relatively little computational time.
%
The computational costs for our methodology are outlined in \fig~\ref{fig:cpu_time}.
%
The computational time is a function of the number of images in the data set, the number of pixels in each point, and the angular resolution to compute the pairwise rotations (see \SI). 
%
Furthermore, the computation of the pairwise rotational alignments, which accounts for the majority of the computational time, is trivially parallelizable for larger data sets.
%
Becasuse the computational cost increases with the image resolution, we chose to subsample all of our data sets to $100 \times 100$ pixels.
%
This resolution allowed us to rapidly analyze our data sets while still capturing all of the relevant developmental features.
%
However, as can be seen from the computational costs in \fig~\ref{fig:cpu_time}, it is feasible to use our algorithms to analyze higher-resolution images.

The requisite user intervention and parameter tuning required for our method is relatively minimal.
%
As a first step, images must be preprocessed so that the Euclidean distance between the pixels is informative.
%
Our software provides several preprocessing options (such as blurring, rescaling, and mean-centering), as well as some guidance for what options to select depending on the system of interest. 
%
Two algorithmic parameters, the angular discretization to compute the pairwise alignments and the diffusion maps kernel scale which determines which data points are ``close'' (see ~\fig~\ref{fig:schematics}A and \SI), must also be defined; we also provide some guidance on selecting these parameters.
%
We found that the results are robust to both of these parameters, and that consistent parameter values served well for all data sets presented here. 
%
Overall, the tasks of image preprocessing and parameter selection are relatively simple compared to manual registration and ordering of images, and so this methodology is promising for much larger imaging data sets which are impractical to evaluate manually.

@@

Already at this point, the rate of collection of imaging data has surpassed the rate of manual image analysis.
%
This necessitates the development of automated methodologies to organize such large data sets. 
%
To the best of our knowledge, algorithmic approaches to these two tasks have been explored largely independently of each other. 
%
Temporal ordering of large-scale data which did not require registration was done in the context of molecular profiling studies, in which data are vectors describing the expression levels of different mRNA \citep{anavy2014blind, trapnell2014dynamics, gupta2008extracting}.
%
At the same time, temporal ordering of imaging data sets was done with a significant amount of human supervision and using registered images as a starting point \citep{yuan2014automated, surkova2008characterization}.  
%
%The task of image registration has been widely studied \citep{zitova2003image}, for applications such as face recognition \citep{rowley1998rotation}, medical image registration \citep{hajnal2010medical}, and texture classification \citep{greenspan1994rotation}.
%
In contrast to most of the existing registration approaches which rely on the knowledge of appropriate landmarks in the images \citep{ian1998statistical} (such as the eyes in face recognition applications \citep{zhao2003face}), algorithms based on angular synchronization can register images even in the absence of such information, making them relevant for a wide variety of applications. 
%
We have applied our methodology to several different proprotypical data sets in developmental biology; because our methodology only relies on computing distances between images and may be readily extended to many other data sets.

We acknowledge that our methods, though general, do have some limitations. 
%
The first is that we require enough amount of data to sufficiently sample the developmental trajectory.
%
Therefore, for very small and/or very noisy data sets, our algorithms may fail. 
%
Second, the developmental features need to be large compared to the noise and the image resolution. 
%
In all of our examples, the relevant features, such as expression patterns and morphological structures, all span several pixels and are large compared to both the instrument noise and embryo-to-embryo variability. 
%
This means that the Euclidean distance is a good measure of similarity between images.
%
For other imaging data sets which contain very fine scale structures, our algorithms will fail as small perturbations in these fine-scale structures will result in large changes in the Euclidean distances. 
%
Furthermore, as we have acknowledged, our methods register full three-dimensional data with respect to rotations due to computational limitations. 
%
However, the algorithms presented here could be used for three-dimensional given an alternative method for computing pairwise alignments more efficiently (such as using image features or landmarks).
%
Unlike traditional statistical methods, our algorithms do not provide confidence intervals on predicted orderings, but could in principle be extracted from bootstrapping \citep{efron1979bootstrap}.
%
Furthermore, although our methods provide a temporal ordering for images, the provide no real-time information, neither do they provide a direction for the proposed ordering.
%
Determining which ``end'' of the trajectory is early and which is late is a post-processing task which requires {\em a priori} knowledge of the developmental dynamics. 
%
Furthermore, real-time information can only be extracted from either {\em a priori} knowledge of the rates of relevant developmental processes, or by making some assumption about the temporal sampling of the data set (e.g., the data set uniformly samples some given time window).

>>>>

\raisebox{4.25cm}{\textsf{C}}
\includegraphics[width=5cm]{drosophila_live_imaging_angle_corr}
%
%\hfill
%
\raisebox{4.25cm}{\textsf{D}}
\includegraphics[width=5cm]{drosophila_live_imaging_rank_corr}
%
%\hfill
%
\raisebox{4.25cm}{\textsf{E}}
%
\raisebox{2.5cm}{
{ \textsf{ 
\begin{tabular}{ |c | c| c| }
\hline
\parbox[c]{1.5cm}{ \centering Movie Index} & \parbox[c]{1.5cm}{ \vspace{.5\baselineskip} \centering Average Angle Error \vspace{.5\baselineskip}} & \parbox[c]{1.5cm}{ \centering Rank Correlation Coefficient} \\
\hline
1 & 8.37$^{\circ}$ & 0.9989 \\ 
2 & 4.98$^{\circ}$ & 0.9994 \\ 
3 & 6.47$^{\circ}$ & 0.9795 \\ 
4 & 7.03$^{\circ}$ & 0.9953 \\ 
5 & 3.04$^{\circ}$ & 0.9959 \\ 
 \hline
\end{tabular}}}}
\caption{Method validation using live imaging of \textit{Drosophila} embryos. (A) Selected images from a live imaging study of a \textit{Drosophilla} embryo during gastrulation. Each frame is in an arbitrary rotational orientation, and the order of the frames has been shuffled. (B) Images from A after registration and ordering using vector diffusion maps. The dorsal side of each embryo now appears at the top of each image, and the ventral side appears at the bottom. (C) The correlation between the recovered rotation angle (using vector diffusion maps) and the true rotation angle. The average absolute error in the recovered angles is $8.37 ^\circ$. (D) The correlation between the recovered rank (using vector diffusion maps) and the true rank. The rank correlation coefficient is $0.9989$. (E) The average error in the recovered angle and the rank correlation coefficient for $5$ different live imaging studies. }
\label{fig:drosophila_live_imaging}
\end{figure*}

@@

We used vector diffusion maps to register and order the images. 
%
\fig~\ref{fig:drosophila_live_imaging}B shows the images from \fig~\ref{fig:drosophila_live_imaging}A, now registered and ordered; the real time for each frame is also indicated.
%
With a small number of exceptions, the recovered ordering is consistent with the real time dynamics. 
%
\fig~\ref{fig:drosophila_live_imaging}C and \fig~\ref{fig:drosophila_live_imaging}D  show the correlations between the recovered and true angles and ranks, respectively, for the entire data set. 
%
Both the angles and the ranks are recovered with a high degree of accuracy.

@@

In general, to accurately recover the temporal order, the vector diffusion maps algorithm requires a sufficient amount of data so that the underlying one-dimensional curve (i.e., the curve in \fig~\ref{fig:schematics}C) is well-sampled.
%
At the same time, the accuracy of the recovered alignments is a function of the signal-to-noise ratio in the pairwise alignments, as well as the size of the data set \citep{singer2011angular}.
%
%Clearly, $40$ images are sufficient to accurately recover the rotational orientation and temporal order in this specific developmental time frame.
%%
%To quantitatively assess how our methods perform as a function of sample size, we used bootstrap sampling to compute the rank correlation as a function of the size of the data set.
%%
%Subsets of images were randomly selected from the $40$-image data set, and then registered and ordered. 
%%
%\fig~\ref{subfig:movie_bootstrap} shows the median rank correlation coefficient as a function of the number of images in the subset.
%%
%As expected, the accuracy in the recovered ordering increases with the number of images. 
%% 
%At the same time, we see that correct order is recovered even with a relatively small number of snapshots, which should make our approach applicable other studies of developmental processes.
%%
Because our particular data set contains relatively low noise levels, few images were required to obtain accurate rotational alignment, and the high temporal resolution of our live imaging study allows us to order the images accurately.  
%
In summary, the algorithm accurately and robustly registers and orders a model data set where both the rotation angles and real times are independently known.

@@

\vspace{0cm}
\raisebox{2.5cm}{\textsf{C}}
%
\includegraphics[width=3.25cm]{zebrafish_rank_corr}
\end{minipage}
\caption{Method validation using live imaging of a zebrafish embryo. (A) Selected images from a flip book movie of zebrafish embryogenesis. Each frame is in an arbitrary rotational orientation, and the order of the frames has been shuffled. (B) Images from A after registration and ordering using vector diffusion maps. The real time of each frame is also indicated. (C) Correlation between the rank recovered using vector diffusion maps and the true rank.  The rank correlation coefficient is $0.9954$. The larger errors in the recovered ranks towards the beginning of the trajectory are due to the rather static cell movement within that time window. }
\label{fig:zebrafish}
\end{figure*}

To illustrate the broad applicability of our approach, we also applied our algorithm to a time-lapse movie  of zebrafish embryogenesis. 
%
We used a publicly available flipbook movie of zebrafish embryogenesis, available at \texttt{https://zfin.org/zf\_info/movies/Zebrafish.mov} \citep{karlstrom1996flipbook}.  
%
Taken with a differential interference contrast (DIC) microscope, the movie records the first $17$~hours of zebrafish development, from a single cell stage to a 16-somite stage. 
%
We selected $120$ consecutive frames from this movie which capture 5.5~hours of epiboly (3.5--9~hours after fertilization). 
%
At the start of the time window, cells have divided 10--11 times and are accumulated in a cell mass above the yolk. 
%
The cell mass is then compressed and the animal-vegetal axis of the embryo (vertical axis in the \fig~\ref{fig:zebrafish}) shortens to form a spherical embryo shape by the end of the fourth hour of development. 
%
Then, the yolk syncytial layer, which forms the boundary between the yolk and the cell mass, moves upward, forming a dome-shaped structure.  
%
During this stage, the cells rearrange to form a cell layer with uniform thickness (about four cells thick). 
%
With time, this cell layer then spreads over across the yolk and expands toward the vegetal pole. 
%
At the end of epiboly, the blastoderm completely engulfs the yolk.

As in the example of {\em Drosophila} embryo live imaging, the frames were randomly rotated and shuffled (\fig~\ref{fig:zebrafish}A). 
%
We then used vector diffusion maps to register and order the frames. 
%
The results are shown in \fig~\ref{fig:zebrafish}B. 
%
The recovered rotations and order are consistent with the expected developmental dynamics, as shown in the correlations between the recovered and true ranks (\fig~\ref{fig:zebrafish}C). 
%
Quantitatively, the rank correlation coefficient for this data set is 0.9954, and the average error in the recovered angle is $4.14^\circ$. 
%
Some errors in ordering images of the early embryo result from rather static cell movement during the early developmental stage where cells divide and accumulate above the yolk. 
%
During epiboly, cell movement is more dynamic and the recovered ordering is more consistent with the real dynamics.

In summary, we showed that our algorithm is not limited to a specific set of data. 
%
The algorithm works for different tissue types ({\em Drosophila} and zebrafish), and also for images taken with different imaging methods (fluorescent and phase contrast images).
%
Provided there exist significant dynamics within the data set and that the developmental trajectory is well-sampled, the developmental dynamics can be recovered.

@@

We have analyzed how our algorithm performs on two model data sets where all images come from a single embryo. 
%
In practice, we are interested in cases where image comes from a different embryo, and the largest source of noise in the considered data set arises from embryo-to-embryo variability.
%
To demonstrate that our methods are robust to such noise sources,
we constructed a synthetic time course data set by selecting a random image from one of five {\em Drosophila} live imaging data sets (those data sets used in \fig~\ref{fig:drosophila_live_imaging}) at each time point.
%
The resulting data set is spatially unregistered, scrambled in time, and reflects embryo-to-embryo variability. 
%
The median rank correlation coefficient when ordering such a synthetic time course using our methodology was 0.77, indicating that the algorithm can accurately recover the temporal order even under noisy conditions.

@@

\fig~\ref{fig:drosophila_fixed_images}B shows the selected images in \fig~\ref{fig:drosophila_fixed_images}A, now registered and ordered using vector diffusion maps \citep{singer2012vector}.
%
However, because of interembryo variability, these selected images are not entirely representative of the developmental dynamics across the entire data set.
%
To highlight the developmental dynamics revealed by vector diffusion maps, we computed an average developmental trajectory across the set of registered and ordered images (see \SI). 
%
Each snapshot in the average trajectory is the (weighted) average of a group of successive images from the registered and ordered data set.
%
Averaging successive images removes (some) of the interembryo variability, so that sequential snapshots of this averaged trajectory, shown in \fig~\ref{fig:drosophila_fixed_images}C, serve as a summary of the stereotypic developmental dynamics.

@@

We have demonstrated that our algorithm can recover the underlying developmental dynamics in several different types of imaging data sets, including ones containing interembryo variability. 
%
Thus far, we have only looked at two-dimensional data sets where biological samples are imaged as a single confocal slice or as a phase contrast image. 
%
However, many developmental processes are inherently three-dimensional, and the z-axis often contains interesting dynamics which are independent of the dynamics in the x-y plane. 
%
Researchers often take multiple z-slices to visualize the three-dimensional structure of a sample.
%
With advances in imaging techniques, such three-dimensional data will be more common, and it is important to validate that our algorithm can be applied to such data.

We used an existing three-dimensional data set of fixed {\em Drosophila} wing imaginal discs, which was published in \citet{hamaratoglu2011dpp}.  
%
Imaginal discs are groups of progenitor cells in fly larva that will transform into a specific organ, such as wings, legs, and eyes, in the adult fly during the pupal transformation. 
%
The wing disc is an imaginal disc that turns into a fly wing. 
%
The data set is composed of 46 fixed wing discs whose developmental times range from 72 to 112~hours after fertilization. 
%
Each disc contains 21 z-slices taken at 1~$\mu$m intervals. 
%
The discs were stained with antibodies that recognize Spalt (red), Dad-GFP (green), and Wingless/Patched (gray) (\fig~\ref{fig:wing_disc}A). 
%
The disc grows in time; at long times, the disc becomes too large to fit completely within the frame and only a partial disc is imaged.

Although the theory of vector diffusion maps readily extends to three-dimensional data, calculation of the pairwise alignments rapidly becomes computational prohibitive for three dimensional data, as one must now search over all three-dimensional rotations (which are parameterized by three Euler angles), rather than over two-dimensional rotations (which are parameterized by a {\em single} rotation angle). 
%
However, ordering data via diffusion maps is readily possible for registered three-dimensional data. 
%
Therefore, we can apply our methodology to three-dimensional data by first registering the data using an alternative approach, and then ordering the registered data using diffusion maps. 
%
For the wing disc data, we registered the three-dimensional data by computing the optimal rotations to align the maximum intensity projections of the data using vector diffusion maps.
%
Because the maximum intensity projections are two-dimensional images, this step is no more computationally intensive than the previous examples. 
%
Other options for registering samples include using a specific localized signal (such as Twi in \fig~\ref{fig:drosophila_fixed_images}) as a marker for alignment, or using other {\em a priori} knowledge.
%
We then used diffusion maps to order the registered three-dimensional data. 
%
\fig~\ref{fig:wing_disc}B shows selected images from the data set ordered by diffusion maps. 
%
In the original data set, each disc was assigned to 6 time classes (72--73~hr, 76.5--77.5~hr, 79--80~hr, 89--90~hr, 100--101~hr, and 110.5--111.5~hr after fertilization); these times are indicated in \fig~\ref{fig:wing_disc}B. 
%
In the ordered set, the size of wing disc grows, and the intensity of proteins increases as a function of time. 
%
The rank correlation coefficient based on the time class is 0.9427. 
%
There are some registration errors in the early images; these are primarily due to some wing discs having extra tissue attached to them. 
%
Even with such obstructions, we can order the images with high accuracy. 
%
Furthermore, we can extract a stereotypical developmental trajectory, shown in \fig~\ref{fig:wing_disc}C, by averaging (see \SI).
%
Thus, this result shows that the algorithm can successfully order a three-dimensional data set.

@@

\raisebox{1.75cm}{\textsf{D}}
{\renewcommand{\arraystretch}{1.5}
{ \textsf{ \small 
\begin{tabular}{ | p{2cm} | c | c | c | c | c | c | }
\hline
Data Set & 
Data Type & 
\parbox[c]{2cm}{ \centering Number of Channels} & 
\parbox[c]{2cm}{ \centering \vspace{.5\baselineskip} Number of Images \vspace{.5\baselineskip}} & 
\parbox[c]{2cm}{ \centering Number of Pixels} & 
\parbox[c]{2cm}{ \centering Angular Discretization} & 
\parbox[c]{2cm}{ \centering CPU time}  \\
\hline
{\em Drosophila} live~imaging & 2D & 1 & 40 & 100 $\times$ 100 & 10$^\circ$ & 3.2~sec \\ 
Zebrafish & 2D & 1 & 120 & 100 $\times$ 100 & 10$^\circ$ & 13~sec\\ 
{\em Drosophila} fixed images & 2D & 3 & 120 & 100 $\times$ 100 & 10$^\circ$ & 29~sec  \\ 
Wing disc & 3D & 3 & 46 & 100 $\times$ 100 $\times$ 21 & 10$^\circ$ & 12~sec  \\
 \hline
\end{tabular}}}}
\caption{Computational requirements for the presented methodology. (A) CPU time as a function of the number of images in the data set (for $100 \times 100$ pixel images, and $10^\circ$ angular discretization). Empirically, the CPU time is $\sim \mathcal{O}(n^{1.33})$ in number of images. (B) CPU time as a function of the number of pixels in the images (for $120$ images, and $10^\circ$ angular discretization). Empirically, the CPU time is $\sim \mathcal{O}(n^{1.83})$ in number of pixels. (C) CPU time as a function of the number of rotations (for $120$ images of $100 \times 100$ pixels). Empirically, the CPU time is $\sim \mathcal{O}(n^{-0.77})$ in the angular discretization. (D) The algorithm settings and computational requirements for the data sets analyzed. All times are reported for an Intel Core i7 2.93 GHz processor.}
\label{fig:cpu_time}
\end{figure*}

Manually ordering the images is nontrivial for researchers who are unfamiliar with the relevant developmental progression, and can be tedious and time-consuming for those who are.
%
In contrast, our method requires relatively little computational time.
%
The computational costs for our methodology are outlined in \fig~\ref{fig:cpu_time}.
%
The computational time is a function of the number of images in the data set, the number of pixels in each point, and the angular resolution to compute the pairwise rotations (see \SI). 
%
Furthermore, the computation of the pairwise rotational alignments, which accounts for the majority of the computational time, is trivially parallelizable for larger data sets.
%
Becasuse the computational cost increases with the image resolution, we chose to subsample all of our data sets to $100 \times 100$ pixels.
%
This resolution allowed us to rapidly analyze our data sets while still capturing all of the relevant developmental features.
%
However, as can be seen from the computational costs in \fig~\ref{fig:cpu_time}, it is feasible to use our algorithms to analyze higher-resolution images.

The requisite user intervention and parameter tuning required for our method is relatively minimal.
%
As a first step, images must be preprocessed so that the Euclidean distance between the pixels is informative.
%
Our software provides several preprocessing options (such as blurring, rescaling, and mean-centering), as well as some guidance for what options to select depending on the system of interest. 
%
Two algorithmic parameters, the angular discretization to compute the pairwise alignments and the diffusion maps kernel scale which determines which data points are ``close'' (see ~\fig~\ref{fig:schematics}A and \SI), must also be defined; we also provide some guidance on selecting these parameters.
%
We found that the results are robust to both of these parameters.
%
Overall, the tasks of image preprocessing and parameter selection are relatively simple compared to manual registration and ordering of images, and so this methodology is promising for much larger imaging data sets which are impractical to evaluate manually.

@@

Already at this point, the rate of collection of imaging data has surpassed the rate of manual image analysis.
%
This necessitates the development of automated methodologies to organize such large data sets. 
%
To the best of our knowledge, algorithmic approaches to these two tasks have been explored largely independently of each other. 
%
Temporal ordering of large-scale data which did not require registration was done in the context of molecular profiling studies, in which data are vectors describing the expression levels of different mRNA \citep{anavy2014blind, trapnell2014dynamics, gupta2008extracting}.
%
At the same time, temporal ordering of imaging data sets was done with a significant amount of human supervision and using registered images as a starting point \citep{yuan2014automated, surkova2008characterization}.  
%
%The task of image registration has been widely studied \citep{zitova2003image}, for applications such as face recognition \citep{rowley1998rotation}, medical image registration \citep{hajnal2010medical}, and texture classification \citep{greenspan1994rotation}.
%
In contrast to most of the existing registration approaches which rely on the knowledge of appropriate landmarks in the images \citep{ian1998statistical} (such as the eyes in face recognition applications \citep{zhao2003face}), algorithms based on angular synchronization can register images even in the absence of such information, making them relevant for a wide variety of applications. 
%
We have applied our methodology to several different proprotypical data sets in developmental biology; because our methodology only relies on computing distances between images,, we are confident it may be readily extended to many other data sets.

We acknowledge that our methods, though general, do have some limitations. 
%
The first is that we require enough amount of data to sufficiently sample the developmental trajectory.
%
Therefore, for very small and/or very noisy data sets, our algorithms may fail. 
%
Second, the relevant developmental patterns and structures need to be large compared to the noise and the image resolution. 
%
In all of our examples, the relevant features, such as expression patterns and morphological structures, all span several pixels and are large compared to both the instrument noise and embryo-to-embryo variability. 
%
This means that the Euclidean distance is a good measure of similarity between images.
%
Our algorithms will fail for other imaging data sets which contain very fine scale structures,  as small perturbations in these fine-scale structures will result in large changes in the Euclidean distances. 
%
In such cases, one requires a more sophisticated distance metric, such as the Earth Mover's Distance \citep{...}, or image description, such as scattering transform coefficients \citep{...}, which are insensitive to such perturbations. 
%
Furthermore, as we have acknowledged, our methods as implemented cannot register full three-dimensional data with respect to rotations due to computational limitations. 
%
However, the algorithms presented here could be used for three-dimensional given an alternative method for computing pairwise alignments more efficiently (such as using image features or landmarks).
